---
title: "Hands-on Exercise 1(2): Thematic Mapping and Geovisualisation"
author: "JUNCHEN LI"
date: "August 27, 2026"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
editor: visual
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

# Overview

This hands-on exercise develops a reproducible thematic-mapping workflow with **sf**, **tidyverse**, and **tmap**. I prepare 2024 resident population data, join it to Singapore's Master Plan 2019 subzones, and design choropleth maps that communicate demographic patterns.

The workflow follows [Chapter 2: Thematic Mapping and GeoVisualisation with R](https://r4gdsa.netlify.app/chap02.html), using the same Master Plan 2019 KML boundary and June 2024 population dataset specified in the chapter.

## Learning objectives

By the end of this exercise, I should be able to:

- prepare demographic variables from a detailed population table;
- join aspatial attributes to a spatial boundary layer;
- produce choropleth maps with suitable classification methods and palettes;
- improve map layout, titles, legends, and explanatory elements; and
- compare several related indicators using small multiples.

# Getting started

## Loading the R packages

```{r}
pacman::p_load(
  sf,
  tidyverse,
  rvest,
  tmap,
  knitr
)

tmap_mode("plot")
```

## Importing the data

```{r}
extract_kml_value <- function(description, field_name) {
  attributes <- description |>
    read_html() |>
    html_table(fill = TRUE) |>
    pluck(1)

  names(attributes)[1:2] <- c("field", "value")

  result <- attributes |>
    filter(field == field_name) |>
    pull(value)

  if (length(result) == 0) NA_character_ else result[[1]]
}

mpsz <- st_read(
  "data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml",
  quiet = TRUE
) |>
  mutate(
    SUBZONE_N = map_chr(
      Description,
      ~ extract_kml_value(.x, "SUBZONE_N")
    ),
    PLN_AREA_N = map_chr(
      Description,
      ~ extract_kml_value(.x, "PLN_AREA_N")
    ),
    REGION_N = map_chr(
      Description,
      ~ extract_kml_value(.x, "REGION_N")
    )
  ) |>
  select(SUBZONE_N, PLN_AREA_N, REGION_N) |>
  st_make_valid() |>
  st_transform(3414)

pop2024 <- read_csv(
  "data/aspatial/respopagesextod2024.csv",
  show_col_types = FALSE
)
```

```{r}
tibble(
  data = c("Subzone boundaries", "Population records"),
  observations = c(nrow(mpsz), nrow(pop2024)),
  year = c(2019, 2024)
) |>
  kable()
```

# Preparing the population indicators

The source table records planning area, subzone, five-year age group, sex, type of dwelling, population, and year. Each resident belongs to one age, sex, and dwelling category, so the detailed records can be summed to subzone level.

```{r}
pop2024 |>
  glimpse()
```

## Defining broad age groups

For this exercise, I define:

- **Young:** age 0--24;
- **Economically active:** age 25--64; and
- **Aged:** age 65 and above.

```{r}
young_ages <- c(
  "0_to_4", "5_to_9", "10_to_14", "15_to_19",
  "20_to_24"
)

active_ages <- c(
  "25_to_29", "30_to_34", "35_to_39", "40_to_44",
  "45_to_49", "50_to_54", "55_to_59", "60_to_64"
)

aged_ages <- c(
  "65_to_69", "70_to_74", "75_to_79", "80_to_84",
  "85_to_89", "90_and_over"
)
```

## Aggregating by subzone

```{r}
pop_profile <- pop2024 |>
  mutate(
    SZ = str_to_upper(SZ),
    PA = str_to_upper(PA),
    Pop = replace_na(Pop, 0)
  ) |>
  group_by(PA, SZ) |>
  summarise(
    YOUNG = sum(Pop[AG %in% young_ages]),
    ECONOMY_ACTIVE = sum(Pop[AG %in% active_ages]),
    AGED = sum(Pop[AG %in% aged_ages]),
    TOTAL = sum(Pop),
    .groups = "drop"
  ) |>
  mutate(
    DEPENDENCY = if_else(
      ECONOMY_ACTIVE > 0,
      (YOUNG + AGED) / ECONOMY_ACTIVE * 100,
      NA_real_
    ),
    PCT_YOUNG = YOUNG / TOTAL * 100,
    PCT_AGED = AGED / TOTAL * 100
  )

pop_profile |>
  arrange(desc(TOTAL)) |>
  slice_head(n = 10) |>
  mutate(
    DEPENDENCY = round(DEPENDENCY, 1),
    PCT_YOUNG = round(PCT_YOUNG, 1),
    PCT_AGED = round(PCT_AGED, 1)
  ) |>
  kable(
    caption = "Ten most populous subzones in the 2024 dataset"
  )
```

# Joining population and boundary data

The spatial file and population table both contain 332 subzones. Converting the subzone names to uppercase provides a direct one-to-one match.

```{r}
mpsz_pop <- mpsz |>
  mutate(SUBZONE_N = str_to_upper(SUBZONE_N)) |>
  left_join(
    pop_profile,
    by = c("SUBZONE_N" = "SZ")
  )

join_check <- mpsz_pop |>
  st_drop_geometry() |>
  summarise(
    boundary_subzones = n(),
    matched_population = sum(!is.na(TOTAL)),
    unmatched_population = sum(is.na(TOTAL))
  )

join_check |>
  kable()
```

# Building thematic maps with tmap

## A basic choropleth map

The first map shows the total resident population. A sequential palette is appropriate because the variable ranges from lower to higher values.

```{r}
#| fig-width: 10
#| fig-height: 7

tm_shape(mpsz_pop) +
  tm_polygons(
    fill = "TOTAL",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.blues"
    ),
    fill.legend = tm_legend(title = "Residents"),
    col = "white",
    lwd = 0.25
  ) +
  tm_title("Resident population by subzone, 2024") +
  tm_layout(
    frame = FALSE,
    legend.outside = TRUE
  )
```

## Comparing classification methods

Different classification methods can emphasise different aspects of the same data. The maps below compare equal intervals with Jenks natural breaks.

```{r}
#| fig-width: 12
#| fig-height: 6

equal_map <- tm_shape(mpsz_pop) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 6,
      values = "brewer.oranges"
    ),
    fill.legend = tm_legend(title = "Dependants per 100")
  ) +
  tm_title("Equal intervals") +
  tm_layout(frame = FALSE)

jenks_map <- tm_shape(mpsz_pop) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "jenks",
      n = 6,
      values = "brewer.oranges"
    ),
    fill.legend = tm_legend(title = "Dependants per 100")
  ) +
  tm_title("Jenks natural breaks") +
  tm_layout(frame = FALSE)

tmap_arrange(equal_map, jenks_map, ncol = 2)
```

Quantile classes contain a similar number of subzones per class, equal intervals preserve equal numerical widths, and Jenks breaks seek internally similar groups. The best choice depends on the distribution and the intended message.

## Refining the map layout

```{r}
#| fig-width: 10
#| fig-height: 7

tm_shape(mpsz_pop) +
  tm_polygons(
    fill = "PCT_AGED",
    fill.scale = tm_scale_intervals(
      style = "jenks",
      n = 6,
      values = "brewer.purples"
    ),
    fill.legend = tm_legend(
      title = "Residents aged 65+ (%)"
    ),
    col = "grey85",
    lwd = 0.3
  ) +
  tm_title("Share of residents aged 65 and above, 2024") +
  tm_compass(type = "8star", position = c("right", "top")) +
  tm_scalebar(position = c("right", "bottom")) +
  tm_credits(
    "Source: Singapore Department of Statistics and data.gov.sg",
    position = c("left", "bottom")
  ) +
  tm_layout(
    frame = FALSE,
    legend.outside = TRUE,
    bg.color = "grey98"
  )
```

# Small multiples

Small multiples use a consistent design to compare several indicators. Each map below uses quantile classes so that the spatial pattern of each variable remains visible despite different numerical ranges.

```{r}
#| fig-width: 13
#| fig-height: 5

young_map <- tm_shape(mpsz_pop) +
  tm_polygons(
    fill = "PCT_YOUNG",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.greens"
    ),
    fill.legend = tm_legend(title = "Young (%)")
  ) +
  tm_title("Age 0–19") +
  tm_layout(frame = FALSE)

aged_map <- tm_shape(mpsz_pop) +
  tm_polygons(
    fill = "PCT_AGED",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.purples"
    ),
    fill.legend = tm_legend(title = "Aged (%)")
  ) +
  tm_title("Age 65+") +
  tm_layout(frame = FALSE)

dependency_map <- tm_shape(mpsz_pop) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "quantile",
      values = "brewer.oranges"
    ),
    fill.legend = tm_legend(title = "Dependency ratio")
  ) +
  tm_title("Dependency") +
  tm_layout(frame = FALSE)

tmap_arrange(
  young_map,
  aged_map,
  dependency_map,
  ncol = 3
)
```

# Exploring selected subzones

The following table identifies subzones with high aged shares. Very small populations can create unstable percentages, so I retain subzones with at least 1,000 residents before ranking them.

```{r}
mpsz_pop |>
  st_drop_geometry() |>
  filter(TOTAL >= 1000) |>
  arrange(desc(PCT_AGED)) |>
  select(SUBZONE_N, PLN_AREA_N, TOTAL, PCT_AGED, DEPENDENCY) |>
  slice_head(n = 10) |>
  mutate(
    PCT_AGED = round(PCT_AGED, 1),
    DEPENDENCY = round(DEPENDENCY, 1)
  ) |>
  kable(
    caption = "Subzones with the highest share of residents aged 65+"
  )
```

# Optional interactive map

Changing to `view` mode produces a zoomable and clickable map. This code is shown for further exploration but is not executed during the normal website render.

```{r}
#| eval: false

tmap_mode("view")

tm_shape(mpsz_pop) +
  tm_polygons(
    fill = "DEPENDENCY",
    fill.scale = tm_scale_intervals(
      style = "jenks",
      values = "brewer.oranges"
    ),
    fill.legend = tm_legend(title = "Dependency ratio"),
    popup.vars = c(
      "Subzone" = "SUBZONE_N",
      "Planning area" = "PLN_AREA_N",
      "Population" = "TOTAL",
      "Aged share (%)" = "PCT_AGED"
    )
  )
```

# Reflection

The exercise demonstrates that a thematic map is shaped by choices made before plotting: variable definitions, name matching, treatment of missing values, class intervals, and colour palettes. Classification methods should therefore be selected deliberately and reported clearly rather than treated as purely visual settings.

# References

- [Geospatial Data Science with R, Chapter 2](https://r4gdsa.netlify.app/chap02.html)
- [Master Plan 2019 Subzone Boundary (No Sea) (KML), data.gov.sg](https://data.gov.sg/datasets/d_0900fdcfdb4666fe2a630f37a532fc06/view)
- [Singapore Department of Statistics](https://www.singstat.gov.sg/)
