---
title: "Hands-on Exercise 1(1): Geospatial Data Science with R"
author: "JUNCHEN LI"
date: "August 27, 2026"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
editor: visual
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

# Overview

This hands-on exercise introduces the **sf** data model and a simple spatial-analysis workflow in R. I work with polygon, line, and point data, check coordinate reference systems, convert a non-spatial table into an `sf` object, and perform buffer, intersection, and point-in-polygon operations.

The workflow is adapted from [Chapter 1: Geospatial Data Science with R](https://r4gdsa.netlify.app/chap01.html). The data files used here are stored locally so that the exercise remains reproducible.

## Learning objectives

By the end of this exercise, I should be able to:

- import GeoJSON and CSV files into R;
- recognise polygon, line, and point geometries;
- inspect and transform coordinate reference systems;
- convert longitude and latitude columns into spatial points; and
- apply basic geoprocessing operations with **sf**.

# Getting started

## Loading the R packages

```{r}
pacman::p_load(
  sf,
  tidyverse,
  rvest,
  knitr
)
```

## Data used

| File | Geometry / type | Purpose |
|---|---|---|
| `MP14_SUBZONE_WEB_PL.shp` | Multipolygon | Master Plan 2014 subzone boundaries |
| `CyclingPathGazette.shp` | Multilinestring | Cycling path network |
| `PreSchoolsLocation.kml` | Point | Preschool locations |
| `listings.csv` | Aspatial CSV | Airbnb listings with coordinates |

The three geospatial datasets were obtained from [data.gov.sg](https://data.gov.sg/), while the listings data came from [Inside Airbnb](https://insideairbnb.com/get-the-data/). The current cycling-path and preschool downloads were converted to the shapefile and KML formats used in the chapter.

# Importing geospatial data

## Subzone boundaries

```{r}
mpsz <- st_read(
  dsn = "data/geospatial",
  layer = "MP14_SUBZONE_WEB_PL",
  quiet = TRUE
) |>
  st_make_valid()

mpsz
```

The object contains one feature for each Master Plan 2014 subzone. Its attributes include the subzone, planning-area, and region names.

## Cycling paths

```{r}
cycling_path <- st_read(
  dsn = "data/geospatial",
  layer = "CyclingPathGazette",
  quiet = TRUE
)

cycling_path
```

## Preschool locations

The preschool GeoJSON stores useful attributes inside an HTML description. The helper function below extracts a requested value from that table.

```{r}
extract_description_value <- function(description, field_name) {
  attributes <- description |>
    read_html() |>
    html_table(fill = TRUE) |>
    pluck(1)

  names(attributes)[1:2] <- c("field", "value")

  result <- attributes |>
    filter(field == field_name) |>
    pull(value)

  if (length(result) == 0) NA_character_ else result[[1]]
}

preschool <- st_read(
  "data/geospatial/PreSchoolsLocation.kml",
  quiet = TRUE
) |>
  mutate(
    CENTRE_NAME = map_chr(
      Description,
      ~ extract_description_value(.x, "CENTRE_NAME")
    )
  )

preschool |>
  st_drop_geometry() |>
  select(CENTRE_NAME) |>
  head() |>
  kable()
```

# Examining spatial objects

## Geometry types and coordinate reference systems

```{r}
spatial_summary <- tibble(
  data = c("Subzones", "Cycling paths", "Preschools"),
  features = c(nrow(mpsz), nrow(cycling_path), nrow(preschool)),
  geometry = c(
    as.character(st_geometry_type(mpsz, by_geometry = FALSE)),
    as.character(st_geometry_type(cycling_path, by_geometry = FALSE)),
    as.character(st_geometry_type(preschool, by_geometry = FALSE))
  ),
  epsg = c(
    st_crs(mpsz)$epsg,
    st_crs(cycling_path)$epsg,
    st_crs(preschool)$epsg
  )
)

spatial_summary |>
  kable()
```

All three files use WGS 84 (`EPSG:4326`). This geographic CRS is useful for sharing data, but distance and area calculations should use a projected CRS. I therefore transform the layers to Singapore SVY21 (`EPSG:3414`).

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
cycling_path_svy21 <- st_transform(cycling_path, 3414)
preschool_svy21 <- st_transform(preschool, 3414)

st_crs(mpsz_svy21)
```

## Plotting different geometry layers

```{r}
#| fig-width: 10
#| fig-height: 6

plot(
  st_geometry(mpsz_svy21),
  col = "grey96",
  border = "grey65",
  main = "Subzones, cycling paths, and preschools"
)
plot(
  st_geometry(cycling_path_svy21),
  col = "#2C7FB8",
  lwd = 0.6,
  add = TRUE
)
plot(
  st_geometry(preschool_svy21),
  col = "#D95F0E",
  pch = 20,
  cex = 0.25,
  add = TRUE
)
```

# Converting aspatial data into an sf object

The Airbnb table contains longitude and latitude values. `st_as_sf()` converts these columns into point geometry.

```{r}
listings <- read_csv(
  "data/aspatial/listings.csv",
  show_col_types = FALSE
)

airbnb_sf <- listings |>
  filter(!is.na(longitude), !is.na(latitude)) |>
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs = 4326,
    remove = FALSE
  ) |>
  st_transform(3414)

airbnb_sf |>
  select(name, room_type, price) |>
  head()
```

```{r}
#| fig-width: 9
#| fig-height: 6

plot(
  st_geometry(mpsz_svy21),
  col = "grey97",
  border = "grey75",
  main = "Airbnb listings in Singapore"
)
plot(
  st_geometry(airbnb_sf),
  col = adjustcolor("#7B3294", alpha.f = 0.45),
  pch = 20,
  cex = 0.45,
  add = TRUE
)
```

# Basic geoprocessing

## Five-metre cycling-path buffer in Tampines West

I first select the Tampines West subzone, retain the cycling paths that intersect it, and then build a five-metre buffer. Because SVY21 uses metres, the buffer distance has a clear physical meaning.

```{r}
tampines_west <- mpsz_svy21 |>
  filter(SUBZONE_N == "TAMPINES WEST")

tampines_cycling <- cycling_path_svy21 |>
  st_filter(tampines_west)

cycling_buffer_5m <- tampines_cycling |>
  st_buffer(dist = 5) |>
  st_intersection(tampines_west)

buffer_area_km2 <- cycling_buffer_5m |>
  summarise() |>
  st_area() |>
  as.numeric() / 1e6

tibble(
  cycling_features = nrow(tampines_cycling),
  buffered_area_km2 = round(buffer_area_km2, 3)
) |>
  kable()
```

```{r}
#| fig-width: 8
#| fig-height: 6

plot(
  st_geometry(tampines_west),
  col = "grey95",
  border = "grey40",
  main = "Five-metre buffer around cycling paths in Tampines West"
)
plot(
  st_geometry(cycling_buffer_5m),
  col = adjustcolor("#41B6C4", alpha.f = 0.65),
  border = NA,
  add = TRUE
)
plot(
  st_geometry(tampines_cycling),
  col = "#253494",
  lwd = 1,
  add = TRUE
)
```

## Counting preschools within each subzone

`st_intersects()` returns the matching preschool points for each polygon. The length of each result is the number of preschools in that subzone.

```{r}
mpsz_analysis <- mpsz_svy21 |>
  mutate(
    PRE_SCH_COUNT = lengths(st_intersects(geometry, preschool_svy21)),
    AREA_KM2 = as.numeric(st_area(geometry)) / 1e6,
    PRE_SCH_DENSITY = PRE_SCH_COUNT / AREA_KM2
  )

mpsz_analysis |>
  st_drop_geometry() |>
  select(SUBZONE_N, PLN_AREA_N, PRE_SCH_COUNT, AREA_KM2, PRE_SCH_DENSITY) |>
  slice_max(PRE_SCH_COUNT, n = 10, with_ties = FALSE) |>
  mutate(
    AREA_KM2 = round(AREA_KM2, 2),
    PRE_SCH_DENSITY = round(PRE_SCH_DENSITY, 2)
  ) |>
  kable(
    caption = "Ten subzones with the largest number of preschools"
  )
```

```{r}
#| fig-width: 9
#| fig-height: 6

plot(
  mpsz_analysis["PRE_SCH_COUNT"],
  pal = function(n) hcl.colors(n, "YlOrRd", rev = TRUE),
  main = "Number of preschools by subzone",
  key.pos = 1
)
```

# Reflection

This exercise shows why the CRS is an analytical decision rather than only a data label. WGS 84 is suitable for storing geographic coordinates, whereas SVY21 is more appropriate for Singapore-based distance and area calculations. The same `sf` structure also allows polygon, line, and point layers to be combined with a consistent set of R verbs.

# References

- [Geospatial Data Science with R, Chapter 1](https://r4gdsa.netlify.app/chap01.html)
- [data.gov.sg](https://data.gov.sg/)
- [Inside Airbnb data portal](https://insideairbnb.com/get-the-data/)
