---
title: "Hands-on Exercise 2(1): First-order Spatial Point Pattern Analysis"
author: "JUNCHEN LI"
date: "September 2, 2026"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
editor: visual
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

# Overview

This exercise examines the first-order spatial pattern of childcare services in Singapore. First-order analysis focuses on spatial intensity: whether events are evenly distributed and where their concentration is relatively high.

The workflow is adapted from [Chapter 4: First-order Spatial Point Patterns Analysis Methods](https://r4gdsa.netlify.app/chap04.html). I use the Clark--Evans nearest-neighbour test to assess clustering and kernel density estimation (KDE) to visualise variations in point intensity.

## Learning objectives

By the end of this exercise, I should be able to:

- prepare point and polygon features for spatial point pattern analysis;
- convert `sf` objects into `ppp` and `owin` objects used by **spatstat**;
- apply the Clark--Evans test under Complete Spatial Randomness (CSR);
- construct and interpret kernel density surfaces; and
- compare point intensity within selected planning areas.

# Getting started

## Loading the R packages

```{r}
pacman::p_load(
  sf,
  spatstat,
  terra,
  tmap,
  tidyverse,
  knitr
)

tmap_mode("plot")
```

## Data used

| File | Geometry | Purpose |
|---|---|---|
| `ChildCareServices.geojson` | Point | Locations and attributes of childcare services |
| `MasterPlan2019SubzoneBoundaryNoSea.geojson` | Polygon | Singapore Master Plan 2019 subzone boundaries |

Both files were downloaded from [data.gov.sg](https://data.gov.sg/) and stored locally in this exercise's `data/geospatial` folder.

# Importing and preparing the data

```{r}
mpsz_sf <- st_read(
  "data/geospatial/MasterPlan2019SubzoneBoundaryNoSea.geojson",
  quiet = TRUE
) |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_make_valid() |>
  st_transform(3414)

childcare_sf <- st_read(
  "data/geospatial/ChildCareServices.geojson",
  quiet = TRUE
) |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_transform(3414)
```

Singapore SVY21 (`EPSG:3414`) is used because distances and areas are measured in metres. The remote island groups are excluded to create a more meaningful mainland analysis window.

```{r}
mpsz_clean <- mpsz_sf |>
  filter(
    SUBZONE_N != "SOUTHERN GROUP",
    !PLN_AREA_N %in% c("WESTERN ISLANDS", "NORTH-EASTERN ISLANDS")
  )

sg_boundary <- mpsz_clean |>
  summarise()

childcare_sg <- childcare_sf |>
  st_filter(sg_boundary)

tibble(
  dataset = c("Planning subzones", "Childcare services"),
  features = c(nrow(mpsz_clean), nrow(childcare_sg)),
  epsg = c(st_crs(mpsz_clean)$epsg, st_crs(childcare_sg)$epsg)
) |>
  kable(caption = "Prepared geospatial datasets")
```

## Mapping the point distribution

```{r}
#| fig-width: 10
#| fig-height: 7

tm_shape(sg_boundary) +
  tm_polygons(fill = "grey96", col = "grey55", lwd = 0.5) +
  tm_shape(childcare_sg) +
  tm_dots(fill = "#E45756", size = 0.018, fill_alpha = 0.55) +
  tm_title("Distribution of childcare services in Singapore") +
  tm_layout(frame = FALSE)
```

# Converting the data for spatstat

The observation window defines where the point process could have occurred. I convert the Singapore boundary to an `owin` object and use the childcare coordinates to create a `ppp` object.

```{r}
sg_owin <- as.owin(st_geometry(sg_boundary))
childcare_xy <- st_coordinates(childcare_sg)

childcare_ppp <- ppp(
  x = childcare_xy[, "X"],
  y = childcare_xy[, "Y"],
  window = sg_owin,
  checkdup = FALSE
)

tibble(
  object = c("Observation window", "Point pattern"),
  class = c(class(sg_owin)[1], class(childcare_ppp)[1]),
  count = c(NA_integer_, npoints(childcare_ppp))
) |>
  kable()
```

```{r}
#| fig-width: 9
#| fig-height: 6

plot(
  childcare_ppp,
  main = "Childcare point pattern within the Singapore window",
  cols = "#2A9D8F",
  pch = 20
)
```

# Clark--Evans nearest-neighbour analysis

The Clark--Evans aggregation index compares the observed mean nearest-neighbour distance with the expected distance under CSR:

- $R < 1$ indicates clustering;
- $R \approx 1$ is consistent with randomness; and
- $R > 1$ indicates a regular or dispersed pattern.

The hypotheses are:

- $H_0$: childcare services follow Complete Spatial Randomness;
- $H_1$: childcare services are spatially clustered.

## Analytical test

```{r}
ce_test <- clarkevans.test(
  childcare_ppp,
  correction = "none",
  alternative = "clustered"
)

tibble(
  aggregation_index_R = unname(ce_test$estimate),
  z_statistic = unname(ce_test$statistic),
  p_value = ce_test$p.value
) |>
  mutate(across(everything(), ~ signif(.x, 4))) |>
  kable(caption = "Clark--Evans test for all childcare services")
```

An index below one indicates that nearest neighbours are closer than would be expected under a random point process. The p-value provides evidence for deciding whether the observed clustering is statistically distinguishable from CSR.

## Monte Carlo CSR test

```{r}
#| cache: true

set.seed(1234)

ce_mc <- clarkevans.test(
  childcare_ppp,
  correction = "none",
  alternative = "clustered",
  method = "MonteCarlo",
  nsim = 99
)

tibble(
  simulations = 99,
  aggregation_index_R = unname(ce_mc$estimate),
  p_value = ce_mc$p.value
) |>
  mutate(across(where(is.numeric), ~ signif(.x, 4))) |>
  kable(caption = "Monte Carlo Clark--Evans test")
```

The simulation-based result compares the observed index with 99 random patterns containing the same number of points in the same study window.

# Kernel density estimation

KDE transforms the discrete childcare locations into a continuous intensity surface. Areas with larger values represent higher concentrations of childcare services, not necessarily areas with greater demand.

## Automatic bandwidth selection

```{r}
#| cache: true

kde_sg <- density(
  childcare_ppp,
  sigma = bw.diggle,
  edge = TRUE,
  kernel = "gaussian"
)

selected_bandwidth <- attr(kde_sg, "sigma")

tibble(
  method = "Diggle cross-validation",
  bandwidth_metres = round(selected_bandwidth, 1)
) |>
  kable(caption = "Selected KDE bandwidth")
```

```{r}
#| fig-width: 10
#| fig-height: 7

plot(
  kde_sg,
  main = "Kernel density of childcare services",
  ribargs = list(las = 1)
)
contour(kde_sg, add = TRUE)
plot(sg_owin, add = TRUE, border = "grey30", lwd = 0.8)
```

## Rescaling intensity

Because SVY21 uses metres, the default KDE values are expressed as points per square metre. Multiplying by one million gives a more readable measure of points per square kilometre.

```{r}
kde_sg_km2 <- eval.im(kde_sg * 1e6)

summary(kde_sg_km2)
```

```{r}
#| fig-width: 10
#| fig-height: 7

plot(
  kde_sg_km2,
  main = "Childcare-service intensity per square kilometre",
  ribargs = list(las = 1)
)
plot(sg_owin, add = TRUE, border = "grey25", lwd = 0.8)
```

# Comparing selected planning areas

The helper function below constructs a separate point pattern for any planning area. I use it to compare Choa Chu Kang and Tampines, the two examples highlighted in the chapter.

```{r}
make_area_pattern <- function(area_name) {
  area_boundary <- mpsz_clean |>
    filter(PLN_AREA_N == area_name) |>
    summarise()

  area_points <- childcare_sg |>
    st_filter(area_boundary)

  area_window <- as.owin(st_geometry(area_boundary))
  area_xy <- st_coordinates(area_points)

  area_ppp <- ppp(
    x = area_xy[, "X"],
    y = area_xy[, "Y"],
    window = area_window,
    checkdup = FALSE
  )

  list(
    boundary = area_boundary,
    points = area_points,
    window = area_window,
    ppp = area_ppp
  )
}

cck <- make_area_pattern("CHOA CHU KANG")
tampines <- make_area_pattern("TAMPINES")

tibble(
  planning_area = c("Choa Chu Kang", "Tampines"),
  childcare_services = c(npoints(cck$ppp), npoints(tampines$ppp)),
  area_km2 = c(area.owin(cck$window), area.owin(tampines$window)) / 1e6,
  services_per_km2 = childcare_services / area_km2
) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable(caption = "Childcare intensity in two planning areas")
```

```{r}
#| fig-width: 12
#| fig-height: 6

par(mfrow = c(1, 2))
plot(cck$ppp, main = "Choa Chu Kang", cols = "#0072B2", pch = 20)
plot(tampines$ppp, main = "Tampines", cols = "#D55E00", pch = 20)
par(mfrow = c(1, 1))
```

```{r}
area_tests <- list(
  "Choa Chu Kang" = cck$ppp,
  "Tampines" = tampines$ppp
) |>
  imap_dfr(function(pattern, area_name) {
    result <- clarkevans.test(
      pattern,
      correction = "none",
      alternative = "clustered"
    )

    tibble(
      planning_area = area_name,
      R = unname(result$estimate),
      p_value = result$p.value
    )
  })

area_tests |>
  mutate(across(where(is.numeric), ~ signif(.x, 4))) |>
  kable(caption = "Clark--Evans results by planning area")
```

# Reflection

This exercise shows that point density and point interaction are different ideas. The Clark--Evans test summarises nearest-neighbour spacing, while KDE reveals where intensity is high or low. KDE results are sensitive to the bandwidth and study boundary, so hotspot-looking areas should be interpreted as exploratory evidence rather than as proof of an underlying cause.

# References

- [R for Geospatial Data Science and Analytics, Chapter 4](https://r4gdsa.netlify.app/chap04.html)
- [Child Care Services, data.gov.sg](https://data.gov.sg/datasets/d_5d668e3f544335f8028f546827b773b4/view)
- [Master Plan 2019 Subzone Boundary (No Sea), data.gov.sg](https://data.gov.sg/datasets/d_8594ae9ff96d0c708bc2af633048edfb/view)
