---
title: "In-class-ex06"
author: "JUNCHEN LI"
date: "2026-10-03"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
editor: visual
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

### Package Loading

::: {style="front-size: 1.50em"}
```{r}
pacman::p_load(sf, sfdep, tmap, 
               plotly, tidyverse, 
               Kendall)
```
:::

### Import data

```{r}
hunan <- st_read(
  dsn = "data/geospatial", 
  layer = "Hunan") %>% 
  st_transform(crs = 32650)
```

### Importing attribute table

```{r}
GDPPC <- read_csv(
  "data/aspatial/Hunan_GDPPC.csv")
```

### Creating a Time Series Cube

```{r}
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

```{r}
is_spacetime_cube(GDPPC_st)
```

### Deriving the spatial weights

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wt = st_inverse_distance(nb, 
                             geometry, 
                             scale = 1,
                             alpha = 1),
    .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```

### Computing Gi\*

```{r}
gi_stars <- GDPPC_nb %>% 
  group_by(Year) %>% 
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>% 
  tidyr::unnest(gi_star)
```

### Mann-Kendall Test on Gi

```{r}
cbg <- gi_stars %>% 
  ungroup() %>% 
  filter(County == "Changsha") |> 
  select(County, Year, gi_star)
```

```{r}
ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()
```

### Interactive Mann-Kendall Plot

```{r}
p <- ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()

ggplotly(p)
```

### Printing Mann-Kendall test report

```{r}
cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>% 
  tidyr::unnest_wider(mk)
```

### Mann-Kendall test data.frame

```{r}
ehsa <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
head(ehsa)
```

### Mann-Kendall test data.frame

```{r}
emerging <- ehsa %>% 
  arrange(sl, abs(tau)) %>% 
  slice(1:10)
head(emerging)
```

### Performing Emerging Hotspot Analysis

```{r}
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st, 
  .var = "GDPPC", 
  k = 1, 
  nsim = 99
)
```

### Visualising the distribution of EHSA classes

```{r}
ggplot(data = ehsa,
       aes(x = classification)) +
  geom_bar()
```

### Visualising EHSA

```{r}
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))
```

```{r}
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(fill_alpha = 0.4)
```
