---
title: "In_class_ex04"
author: "JUNCHEN LI"
date: "September 19, 2026"
date-modified: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
editor: visual
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

### Package Loading

::: {style="front-size: 1.50em"}
```{r}
pacman::p_load(
  sf,
  tidyverse,
  rvest,
  knitr,
  spdep,
  tmap,
  GWmodel,
  ggstatsplot
)
```
:::

### data Preparing

```{r}
hunan_sf <- st_read(dsn = "data/geospatial", layer = "Hunan")
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")

hunan_sf <- left_join(hunan_sf, hunan2012) %>%
  select(1:3, 7,15,16,31,32)
```

### Mapping GDPPC

::: panel-tabset
## DIY

```{r, echo=FALSE}
#| fig-width: 16
#| fig-height: 6
basemap <- tm_shape(hunan_sf) + 
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan_sf, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap,gdppc,asp = 1,ncol = 2)
```

## The code

```{r, eval=FALSE}
basemap <- tm_shape(hunan_sf) + 
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan_sf, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap,gdppc,asp = 1,ncol = 2)
```
:::

### Converting to SpatialPolygonDataFrame

```{r}
hunan_sp <- hunan_sf %>%
  as_Spatial()
```

### Geographically Weighted Summary Statistics(GSWW) with adaptivev bandwidth

#### Determine adaptive bandwidth

::: panel-tabset
### Cross-validation

```{r}
bw_CV <- bw.gwr(GDPPC ~1,
                data = hunan_sp,
                approach = "CV",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

### AICc

```{r}
bw_AIC <- bw.gwr(GDPPC ~1,
                data = hunan_sp,
                approach = "AICc",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

```{r}
bw_AIC
```
:::

## GWSS with Fixed Bandwidth

### Determine fixed bandwidth

::: panel-tabset
#### Cross-Validation

```{r}
bw_CV <- bw.gwr(GDPPC ~1,
                data = hunan_sp,
                approach = "CV",
                adaptive = FALSE,
                kernel = "bisquare",
                longlat = T)
```

#### AICc

```{r}
w_AIC <- bw.gwr(GDPPC ~1,
                data = hunan_sp,
                approach = "AICc",
                adaptive = FALSE,
                kernel = "bisquare",
                longlat = T)
```

```{r}
bw_AIC
```
:::

## Computing geographically wieghted summary statistics

```{r}
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               kernel = "bisquare",
               quantile = TRUE,
               adaptive = TRUE,
               longlat = T)
```

## Preparing the output data

```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
```

```{r}
hunan_gstat <- cbind(hunan_sf, gwstat_df)
```


## Visualising geographically weighted summary statistics

### The Geographically Weighted Mean
```{r, echo=FALSE}
#| fig-width: 16
#| fig-height: 6
tm_shape(hunan_gstat) + 
  tm_polygons(fill = "GDPPC_LM", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.oranges"),
              fill.legend = tm_legend(
                title = "local Mean")) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of geographically weighted mean") +
  tm_layout(legend.position = c("left", "bottom"))
```

### The Code
```{r, eval=FALSE}
#| fig-width: 16
#| fig-height: 6
tm_shape(hunan_gstat) + 
  tm_polygons(fill = "GDPPC_LM", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.oranges"),
              fill.legend = tm_legend(
                title = "local Mean")) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of geographically weighted mean") +
  tm_layout(legend.position = c("left", "bottom"))
```

