---
title: "Hands-on Exercise 5(1): Global Spatial Autocorrelation"
author: "JUNCHEN LI"
date: "September 24, 2026"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    code-tools: true
editor: source
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

# Overview

This exercise accompanies [Chapter 9](https://r4gdsa.netlify.app/chap09.html). The question is whether neighbouring Hunan counties have similar GDP per capita in 2012. The analysis treats county boundaries as fixed and tests the arrangement of an attribute, rather than the locations of point events. [Part 2](Hands-on_Ex0502.qmd) examines individual counties.

# 1 Data and preparation

Both parts use `data/geospatial/Hunan.shp` and its companion files, together with `data/aspatial/Hunan_2012.csv`. The data are downloaded from the instructor's [pinned course repository](https://github.com/tskam/ISSS626-AY2026-27Aug/tree/966eea6df498b38f10bc209d07a922f4d87b95d9/lesson/Lesson04/data). `data/download-data.R` retrieves the inputs; `data/download-manifest.txt` records URLs, retrieval date and checksums. **2012 is the indicator year**, not the download year. Values retain the supplied scale.

## Packages

Install missing packages once in the R console before rendering. The document does not install software during a render.

```{r}
#| label: setup
library(sf)
library(spdep)
library(tmap)
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)
tmap_mode("plot")
theme_set(theme_minimal(base_size = 12))
```

## Import and join

An explicit county key prevents an accidental join on unrelated columns. The checks require a unique match for every polygon and retain polygon order for the weights.

```{r}
#| label: import-data
boundary <- st_read("data/geospatial/Hunan.shp", quiet = TRUE)
attributes <- read_csv("data/aspatial/Hunan_2012.csv", show_col_types = FALSE)
stopifnot(nrow(boundary) == 88L, nrow(attributes) == 88L,
          !anyDuplicated(boundary$County), !anyDuplicated(attributes$County),
          setequal(boundary$County, attributes$County),
          all(st_is_valid(boundary)), !any(st_is_empty(boundary)))
hunan <- left_join(boundary, select(attributes, County, GDPPC),
                   by = "County", relationship = "one-to-one")
stopifnot(identical(hunan$County, boundary$County),
          all(is.finite(hunan$GDPPC)), all(hunan$GDPPC > 0))
data.frame(Counties = nrow(hunan), Missing_GDPPC = sum(is.na(hunan$GDPPC)),
           Minimum = min(hunan$GDPPC), Median = median(hunan$GDPPC),
           Maximum = max(hunan$GDPPC)) |> kable()
```

# 2 Two views of the same attribute

```{r}
#| label: fig-classification
#| fig-width: 12
#| fig-height: 7
#| fig-cap: "Equal-interval and quantile classifications of Hunan GDPPC, 2012."
indicator_map <- function(style, title) {
  tm_shape(hunan) +
    tm_polygons(fill = "GDPPC",
      fill.scale = tm_scale_intervals(style = style, n = 5,
                                      values = "brewer.blues"),
      fill.legend = tm_legend(title = "GDPPC (2012)"),
      col = "white", lwd = 0.4) +
    tm_title(title, size = 1) +
    tm_layout(frame = FALSE, legend.outside = TRUE)
}
tmap_arrange(indicator_map("equal", "Equal intervals"),
             indicator_map("quantile", "Quantiles"), ncol = 2)
```

Equal intervals preserve equal numerical widths, while quantiles distribute counties more evenly among classes. A colour cluster depends partly on these choices. Neither map alone supplies a statistical test, and GDP per capita should not be interpreted as every resident's income.

# 3 Queen neighbours and spatial weights

```{r}
#| label: queen-weights
queen <- poly2nb(hunan, queen = TRUE, row.names = hunan$County)
weights <- nb2listw(queen, style = "W", zero.policy = FALSE)
stopifnot(all(card(queen) > 0),
          max(abs(vapply(weights$weights, sum, numeric(1)) - 1)) < 1e-12)
data.frame(Counties = length(queen), Directed_links = sum(card(queen)),
           Minimum_neighbours = min(card(queen)),
           Maximum_neighbours = max(card(queen)),
           Mean_neighbours = mean(card(queen))) |> kable(digits = 3)
```

There are **`r sum(card(queen))` directed links**. A shared boundary point is sufficient for queen adjacency. Row standardisation gives each county a weighted neighbour average; it does not correct for omitted counties outside the province. Changing the neighbour definition changes the question being tested.

# 4 Global Moran's I

## Analytical test

The null randomly assigns the observed GDPPC values to the fixed counties. The one-sided alternative tests positive spatial association. Exchangeability is an assumption, not proof that economic development is generated randomly.

```{r}
#| label: moran-test
moran_result <- moran.test(hunan$GDPPC, weights,
                           alternative = "greater", randomisation = TRUE)
moran_result
```

The observed I is **`r round(unname(moran_result$estimate[1]), 4)`**, compared with the randomisation expectation **`r round(-1/(nrow(hunan)-1), 4)`**. The analytical p-value is **`r format.pval(moran_result$p.value, digits = 3)`**. A positive value indicates a province-wide tendency for similar values to be adjacent; it does not locate individual clusters or identify a cause.

## Monte Carlo test

```{r}
#| label: moran-permutation
set.seed(1234)
moran_sim <- moran.mc(hunan$GDPPC, weights, nsim = 999,
                      alternative = "greater")
moran_sim
```

```{r}
#| label: fig-moran-null
#| fig-cap: "Permutation distribution of Moran's I; the red line marks the observed statistic."
ggplot(data.frame(I = moran_sim$res[seq_len(999)]), aes(I)) +
  geom_histogram(bins = 25, fill = "#4878a8", colour = "white") +
  geom_vline(xintercept = unname(moran_sim$statistic), colour = "#b2182b", linewidth = 1) +
  geom_vline(xintercept = -1/87, linetype = "dashed") +
  labs(x = "Moran's I under random reassignment", y = "Permutations")
```

The simulated reference contains only the 999 reassigned datasets; the observed value is excluded from the histogram. The permutation p-value is **`r moran_sim$p.value`**. With 999 permutations, tail probabilities have limited resolution. This test and the analytical test use the same data and weights, so agreement is not independent replication.

## Moran scatterplot

```{r}
#| label: fig-moran-scatter
#| fig-cap: "Standardised GDPPC and its spatial lag using row-standardised queen weights."
z <- as.numeric(scale(hunan$GDPPC))
wz <- lag.listw(weights, z)
slope <- unname(coef(lm(wz ~ 0 + z))[1])
stopifnot(abs(slope - unname(moran_result$estimate[1])) < 1e-10)
ggplot(data.frame(z, wz), aes(z, wz)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(colour = "#2166ac", size = 2) +
  geom_abline(intercept = 0, slope = slope, colour = "#b2182b") +
  labs(x = "Standardised GDPPC", y = "Spatial lag of standardised GDPPC")
```

The line through the origin has slope I because the weights sum to the number of counties. The lag is computed **after** standardising GDPPC; independently standardising the lag would change this slope relationship. Quadrants describe relative values, not local significance.

# 5 Geary's C

```{r}
#| label: geary-tests
geary_result <- geary.test(hunan$GDPPC, weights,
                           alternative = "greater", randomisation = TRUE)
set.seed(1234)
geary_sim <- geary.mc(hunan$GDPPC, weights, nsim = 999,
                      alternative = "greater")
geary_result
geary_sim
```

Geary's C uses squared differences between neighbours. Its null reference is one; smaller values indicate positive association. Here C is **`r round(unname(geary_result$estimate[1]), 4)`** with permutation p = **`r geary_sim$p.value`**. In these spdep functions, `alternative = "greater"` means greater positive autocorrelation, corresponding to the **lower** tail of C.

```{r}
#| label: fig-geary-null
#| fig-cap: "Permutation distribution of Geary's C; the observed value is red and the null reference of one is dashed."
ggplot(data.frame(C = geary_sim$res[seq_len(999)]), aes(C)) +
  geom_histogram(bins = 25, fill = "#4d9d88", colour = "white") +
  geom_vline(xintercept = unname(geary_sim$statistic), colour = "#b2182b", linewidth = 1) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(x = "Geary's C under random reassignment", y = "Permutations")
```

# 6 Spatial correlograms

A lag here is a step in the county neighbour graph, not a kilometre distance or a time interval. Higher-order neighbours are defined by their shortest graph path.

```{r}
#| label: correlograms
mi_corr <- sp.correlogram(queen, hunan$GDPPC, order = 6,
                         method = "I", style = "W", zero.policy = TRUE)
gc_corr <- sp.correlogram(queen, hunan$GDPPC, order = 6,
                         method = "C", style = "W", zero.policy = TRUE)
print(mi_corr, p.adj.method = "holm")
print(gc_corr, p.adj.method = "holm")
```

```{r}
#| label: fig-correlograms
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Moran and Geary summaries across six graph lags."
par(mfrow = c(1, 2))
plot(mi_corr, main = "Moran's I")
plot(gc_corr, main = "Geary's C")
par(mfrow = c(1, 1))
```

Read the estimates together with the printed tests. Holm adjustment is applied separately to the six lags for each statistic; it is an exploratory check, not a correction for every decision in this exercise. Associations at wider graph lags can differ from immediate-neighbour association. The plot's uncertainty bars are not simultaneous confidence bands.

# 7 Interpretation and limitations

The main queen-weight tests give I = **`r round(unname(moran_sim$statistic), 3)`** and C = **`r round(unname(geary_sim$statistic), 3)`**, both consistent with positive spatial association in this dataset. The result describes relative county GDPPC patterns in **2012**, not changes through time. The analysis cannot establish spillover effects, causal mechanisms or local significance. County aggregation, unusually large indicator values and the province boundary can affect the summaries. Local analysis follows in [Exercise 5(2)](Hands-on_Ex0502.qmd).

# References and session information

- [Chapter 9: Global Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap09.html).
- [spdep: Moran's I](https://r-spatial.github.io/spdep/reference/moran.test.html), [Geary's C](https://r-spatial.github.io/spdep/reference/geary.test.html), and [correlograms](https://r-spatial.github.io/spdep/reference/sp.correlogram.html).

```{r}
#| label: session-info
sessionInfo()
```
